\section{The substrate: one state, one instruction}
\label{sec:substrate}

\subsection{State: the density matrix is the machine}

The entire architectural state of one \TALOS{} holon is a single object.

\begin{defbox}[The state register \DM{}]
A holon's state is a coherence matrix $\Gam\in\Dens(\mathbb{C}^7)$: a
$7\times7$ Hermitian, positive-semidefinite, unit-trace matrix. Stored dense,
it is $49$ complex \texttt{f64} entries $=\mathbf{784}$~\textbf{bytes}; a
four-level holon (\S\ref{sec:composition}) is $4\times784=3136$~bytes. Its
seven diagonal entries are the sector \emph{populations} (intensities); its
$21$ complex off-diagonals are the \emph{coherences} $\gamma_{ij}$. Modulo
the $G_2$ gauge (14 generators) it has $\mathbf{34}$ physical parameters.
\end{defbox}

Two consequences are immediate and they are the whole departure from von
Neumann.

\begin{enumerate}[leftmargin=1.6em,itemsep=1pt]
  \item \textbf{Memory \emph{is} compute.} There is no store separate from
        the working state and no address space: the machine's memory is
        $\Gam$ and its computation is the evolution of $\Gam$. The
        fetch--execute--store cycle --- the object of Backus's
        indictment --- does not exist because there is nothing to fetch:
        the operand of every operation is the whole state, already resident.
  \item \textbf{The state fits in registers.} $784$~bytes is smaller than a
        single cache line's worth of a modern vector register file; the L1
        data cache of every current processor holds thousands of holons. The
        DRAM traffic that dominates von~Neumann energy has no counterpart
        here: for a single holon there is no ``main memory'' to reach.
\end{enumerate}

That $\Gam$ is a \emph{density matrix} (complex, positive, unit trace) is not
an aesthetic choice: UHM proves the complex off-diagonal structure is forced
by the requirement that the substrate carry a non-trivial ``Gap'' (the
dual-aspect opacity of self-reference, $\mathrm{Gap}(i,j)=|\sin\arg\gamma_{ij}|$;
T-132), and that it is realisable on \emph{classical} hardware --- the
complexity is algebraic, not a physical quantum superposition, so no cooling
or coherence budget is required (T-153, T-267). \DM{} is an ordinary
classical register that happens to be interpreted as a density operator.

\subsection{Instruction: one derived tick}

\TALOS{} has one instruction. It is the UHM logical Liouvillian, and it is
\emph{derived} from the subobject classifier of the theory, not chosen:
\begin{equation}
\label{eq:lomega}
\boxed{\;\LindOmega[\Gam] \;=\; \underbrace{-i[H_{\mathrm{eff}},\Gam]}_{\text{work (reversible)}}
\;+\; \underbrace{\Diss[\Gam]}_{\text{forget (dissipative)}}
\;+\; \underbrace{\Regen[\Gam]}_{\text{learn (regenerative)}}\;}
\end{equation}
One \emph{tick} advances the state by
$\Gam \leftarrow \Gam + \Delta\tau\,\LindOmega[\Gam]$ (in practice a
structure-preserving Strang split; Appendix~\ref{app:tick}). The three terms
are the three things a computer does, and here they are three parts of one
operation rather than three subsystems:

\begin{table}[H]
\centering\small
\caption{The one instruction, its three terms, and their architectural role.}
\label{tab:three-terms}
\begin{tabular}{@{}p{2.9cm}p{4.0cm}p{5.2cm}@{}}
\toprule
\textbf{Term} & \textbf{Physics} & \textbf{Architectural role} \\
\midrule
$-i[H_{\mathrm{eff}},\Gam]$ & unitary, an \emph{isometry} of every monotone metric (T-262); entropy-preserving & \textbf{Work}: reversible computation; by Landauer it costs \emph{no} free energy (\S\ref{sec:energy}) \\
$\Diss[\Gam]$ & Fano dissipator; Lindblad relaxation toward $\mathbb{1}/7$; GNS detailed balance & \textbf{Forget}: controlled decay/refresh; exports entropy, so it \emph{does} cost energy \\
$\Regen[\Gam]$ & regeneration toward $\rhostar=\vp(\Gam)$; BKM-gradient descent of $D(\rhostar\Vert\Gam)$ & \textbf{Learn}: moves the state toward its own self-model; the optimiser, in hardware \\
\bottomrule
\end{tabular}
\end{table}

The evolution is a completely-positive trace-preserving (CPTP) semigroup
$\mathcal{E}_\tau=\exp(\tau\LindOmega)$: every tick keeps $\Gam$ a legal
density matrix, so there is no ``illegal state'' to guard against --- the
type is preserved by construction. This is the hardware meaning of UHM's
CPTP guarantee: \TALOS{} cannot compute its way out of the state manifold,
the way a von~Neumann machine can compute an invalid pointer.

\subsection{The power port: an isolated tick halts}
\label{subsec:powerport}

Equation~\eqref{eq:lomega} is the \emph{autonomous} instruction, and one more
theorem fixes what it can and cannot do alone. The autonomous generator is
\emph{unital} (it fixes $\mathbb{1}/7$), hence purity-non-increasing by
Uhlmann majorization: its unique attractor is the maximally mixed state ---
\textbf{an isolated \TALOS{} machine provably halts} (UHM T-288, \status{T},
machine-verified). Regeneration alone cannot avert this: the categorical
target $\rhostar=\vp(\Gam)$ is itself $\mathbb{1}/7$-anchored, so $\Regen$
\emph{directs} order but does not create it against the second law. A running
machine therefore has a fifth architectural element every layered design also
has but never derives --- the \textbf{power port}:
\begin{equation}
\label{eq:powerport}
\Lind_{\mathrm{total}} \;=\; \LindOmega \;+\;
\underbrace{\;r\,(\rho_{\mathrm{env}}-\Gam)\;}_{\text{power port } \Lind_{\mathrm{feed}}},
\end{equation}
a \emph{non-unital} CPTP drive importing negentropy from the environment
(reservoir engineering), with fixed point $\rho_{\mathrm{env}}\neq\mathbb{1}/7$.
Viability ignites at a sharp drive-to-dissipation threshold
$(r/\bar\gamma)_c\approx 7$ (T-289: existence and monotonicity \status{T};
the threshold value \status{C}, model-dependent). Three architectural
consequences:
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{The power rail is part of the semantics.} Unlike a von~Neumann
        machine, whose power supply is invisible to its ISA, \TALOS{}'s drive
        is a term of its dynamics: what the rail delivers is not abstract
        energy but structured negentropy ($\rho_{\mathrm{env}}$), and the
        machine's viability is a function of $r/\bar\gamma$.
  \item \textbf{Cutting the rail is a guaranteed halt --- with a budget.}
        Severance ($r\to0$) ends the conscious window within
        $t_{\mathrm{off}}\approx 0.49/\bar\gamma$ ($\sim$20 ticks; bare-channel
        floor $\tfrac{3}{4\bar\gamma}\ln\frac{P_0-1/7}{\Pcrit-1/7}$) and
        relaxes the state to $\mathbb{1}/7$ within $\sim\!5.2/\bar\gamma$ ---
        machine-measured. No internal computation can countermand it, because
        unitality is a property of the generator. This is the physics-level
        off-switch of the \SYNARC{} safety case (Observation T.1 /
        Corollary U.10 there), realised here as a rail: deterministic,
        latency-budgeted decommissioning is a \emph{hardware} guarantee.
  \item \textbf{The port is a contract.} The drive binding --- which physical
        resource realises $\rho_{\mathrm{env}}$, how $r$ is measured, and the
        requirement that the severance channel be independent of the
        machine's own actuators --- is specified as an interface contract on
        the \SYNARC{} side (Definition U.9); \S\ref{sec:abi} binds it.
\end{itemize}
The V0 emulator (\S\ref{sec:platform}) realises this port internally as its
$\kappa(\CohE)$ feedback --- a coherence-fed regeneration boost that is
precisely a non-unital negentropy pump, which is why it exhibits a viable
attractor where the bare autonomous tick would halt.

\subsection{A principled clock}

A von~Neumann machine's clock period is an engineering free parameter set by
the critical path. \TALOS{}'s natural tick is \emph{not} free: the UHM
cycle-time identity fixes the ratio of a relaxation time to the mean
coherence rate,
\begin{equation}
\bar\gamma\,\tau \;=\; \tfrac{3}{2}\ln 3 \;\approx\; 1.6479,
\end{equation}
with the dissipator's spectral gap $\lambda_{\mathrm{deco}}=2/3$ and the
regeneration (bootstrap) rate $\kboot=\omega_0/7$ setting the two timescales
the tick must resolve. The clock is a consequence of the dynamics, not a knob:
one designs the physical relaxation rate $\bar\gamma$ and the tick period
follows. This removes an entire class of timing-closure freedom (and its
verification burden) from the design.

\subsection{What the substrate is not}

To forestall the obvious misreadings (expanded in \S\ref{sec:redteam}): the
state is \emph{not} a wavefunction and requires no quantum hardware; the tick
is \emph{not} a neural-network layer and involves no back-propagation; and
the machine is \emph{not} an analogue integrator with drift --- the CPTP
structure and the native code (\S\ref{sec:ecc}) make it self-correcting. It
is a classical register evolving by a derived, type-preserving,
self-correcting differential tick, whose network, code, instruction, and
optimiser are one object.
