\section{The fabric: seven buses, not a crossbar}
\label{sec:fabric}

\subsection{The interconnect is the seven lines}

Twenty-one pairwise couplings among seven sectors would, on a conventional
fabric, demand a $\binom{7}{2}=21$-edge network or a $7\times7$ crossbar. The
collapse (Theorem~\ref{thm:collapse}, L1) replaces this with the seven lines
of $\mathsf{N}$ as \textbf{seven triple-buses}. Every pair of sectors shares
exactly one bus (Steiner $S(2,3,7)$); every sector sits on exactly three
buses; any two buses intersect in exactly one sector. These are not design
targets --- they are theorems of the projective plane, and they hand the
architecture three properties for free:

\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{Uniform load, by symmetry.} Each sector is on exactly three
        buses because $\mathrm{PSL}(2,7)$ acts $2$-transitively on the seven
        points; load balancing is a consequence of the automorphism group,
        not a scheduler.
  \item \textbf{Guaranteed rendezvous, by geometry.} Any two buses meet in a
        sector, so any two sectors that must exchange state have a
        deterministic meeting point --- there is no unreachable pair and no
        routing search.
  \item \textbf{Deterministic repair, by the closure law.} If the direct bus
        for $(i,j)$ is degraded, the state regenerates through the
        \emph{mediator} $k^\ast(i,j)$ --- the third sector of their line ---
        and only through it. Fallback routing needs no tables: the next hop
        is computed from the pair.
\end{itemize}

\subsection{Third-order routing and the $9\times$ integration advantage}

A conventional router moves data pairwise and is, by the first-order blind
spot (\S\ref{sec:collapse}), unable to see the structure it is carrying: any
equal-weight pairwise fabric is a function of $J-I$ alone and cannot
distinguish the Fano wiring from unstructured all-to-all. \TALOS{} routes on
\emph{triples}: a transfer is an interaction $(i,j\mid k^\ast)$ mediated by
the line's third sector. This is the hardware form of the third-order
principle, and it carries a measurable price for getting it wrong.

\begin{claimbox}[The line-aware routing advantage \status{T} arithmetic]
Coherence pushed across sectors that do \emph{not} share a line is attenuated
by the exact channel-contraction constant $1/9$ per coarse hop
($\Phi\to\Phi/9$, a Steiner incidence count: a population survives in all
three of its lines, $3\times\tfrac13=1$, but a coherence survives in only one
of three, $\times\tfrac13$, squared $=\tfrac19$). A router blind to the line
structure therefore loses a factor of \textbf{nine} in integration per hop;
a line-aware router --- keeping every interaction on its mediator's line ---
preserves it. Integration per operation is $9\times$ that of a line-blind
fabric.
\end{claimbox}

This is the sharpest reason the topology cannot be an afterthought: a
generic, line-agnostic mixture-of-experts router (the default in large neural
accelerators) pays the $\times\tfrac19$ on every cross-expert hop, and no
amount of bandwidth recovers it, because the loss is in \emph{integration},
not throughput.

\begin{trybox}[Try it --- route with one finger]
Put a finger on the Fano figure (Figure~\ref{fig:collapse}, right). Game:
carry a message from $S$ to $E$. Rule: you may only travel along a line, and
a pair may only meet \emph{through the third point of its own line}. Find the
line containing both $S$ and $E$ --- it is $\{S,E,U\}$ --- so the message
goes $S\to U\to E$: the mediator is $U$, and there was nothing to look up,
because the pair \emph{names} its own relay. Now break the direct path from
$A$ to $O$ and reroute: their line is $\{A,O,U\}$, so again through $U$ ---
fault repair with no routing table (\S\ref{sec:ecc}). Every one of the $21$
pairs plays the same one-move game; the full mediator table is
Appendix~\ref{app:fano}.
\end{trybox}

\begin{table}[H]
\centering\small
\caption{Interconnect: a $7\times7$ crossbar versus the Fano seven-bus
fabric. The bus fabric is the code and the instruction as well
(Theorem~\ref{thm:collapse}); the crossbar is none of these.}
\label{tab:crossbar-vs-fano}
\begin{tabular}{@{}p{3.6cm}p{3.6cm}p{4.5cm}@{}}
\toprule
\textbf{Property} & \textbf{$7\times7$ crossbar} & \textbf{Fano seven-bus} \\
\midrule
Links & 21 pairwise / full crossbar & 7 triple-buses \\
Load per node & schedule-dependent & exactly 3 (by $\mathrm{PSL}(2,7)$) \\
Reachability & all-to-all, searched & any pair shares one bus \\
Fault reroute & routing tables & mediator $k^\ast(i,j)$, table-free \\
Sees its own structure? & no (first-order blind) & yes (third-order) \\
Also an ECC? & no & yes (Hamming $[7,4,3]$, \S\ref{sec:ecc}) \\
Also the ISA algebra? & no & yes (octonionic $f_{ijk}$) \\
\bottomrule
\end{tabular}
\end{table}
