\section{Error correction is the topology}
\label{sec:ecc}

\subsection{The wiring is the code}

In a conventional machine, fault tolerance is a separate layer: ECC-DRAM
dedicates check bits (a $+12.5\%$ storage overhead for the common
$[72,64]$ SEC-DED code) and a syndrome decoder; a quantum machine spends on
the order of $10^3$ physical qubits per logical qubit on a surface code.
The overhead exists because the code is a structure \emph{added to} the
computation. \TALOS{} adds nothing, because the interconnect already
\emph{is} a distance-3 code.

\begin{claimbox}[Zero-overhead fault localisation \status{T}]
The Fano incidence is the parity-check matrix of the Hamming code
$[n,k,d]=[7,4,3]$: single errors are corrected, double errors detected, and a
single degraded sector is \emph{located} by three syndrome checks --- the
diagnosis pyramid $21\to 7\to 3\to 1$ (twenty-one pair-relations collapse to
seven axes, read out by three syndrome bits, yielding one fault verdict). The
three parity checks are the three lines through the faulty point. No storage,
no decoder array, and no code-versus-computation mismatch: the fabric of
\S\ref{sec:fabric} \emph{is} the check matrix.
\end{claimbox}

The syndrome map is exact and was machine-verified: for each of the seven
single-sector fault patterns, the three-bit syndrome read as a binary index
returns the fault position (Appendix~\ref{app:fano}). Because the code is the
Hamming code, its guarantees are the textbook ones ($d=3$: correct 1, detect
2), inherited at zero marginal cost.

\begin{zerobox}[From zero --- find the broken toy with three questions]
Seven toys stand in a row, numbered $1$ to $7$. Exactly one is broken, and
you may ask only yes/no questions. Three questions suffice --- \emph{if} you
ask the right ones: ``is its number $4$ or more?'', ``is it in
$\{2,3,6,7\}$?'', ``is it odd?''. Write the three answers as $1$s and $0$s
and you have spelled the broken toy's number in binary: three answers, seven
suspects, one verdict. The Fano wiring asks exactly these three questions
\emph{automatically, on every tick} --- each question is one of the three
lines through the faulty sector --- and that is the whole error-correcting
code: $21\to7\to3\to1$, no extra parts, because the wiring \emph{is} the
questionnaire.
\end{zerobox}

\begin{mythbox}[The myth, continued --- he knows where it hurts]
Strike the giant anywhere and he does not search his body for the wound: the
same bronze that moves him reports the place. Talos needs no physician
walking behind him with a checklist --- and neither does this machine
(Principle~\ref{prin:selfdiag}): the tick that computes is the tick that
diagnoses, because the pattern that carries the state is the pattern that
checks it.
\end{mythbox}

\paragraph{The rectangle invariant: a second, deeper parity layer (T-281/T-282).}
The Hamming layer above guards the \emph{axes} (which sector is faulty). The
corpus's genesis results add a guard one level down: the \emph{sign firmware}
of the fabric itself. Viability of the Fano sign table --- ``no two states can
annihilate'' --- is provably a \emph{linear} condition: the 84 rectangle rules
$F(p,r)F(q,s)F(p,s)F(q,r) = -1$ over the XOR-classes, whose exhaustive solution
set is exactly one gauge orbit of the octonions (corpus T-281). A bit-flip in
the stored sign table therefore either lands inside the 16-element gauge orbit
(harmless relabeling) or violates at least one rectangle --- an \emph{84-bit
XOR self-test of the firmware}, as cheap as a parity sweep, that certifies the
machine's multiplication law is still the viable one. And the same algebra
explains why no firmware patch can ever ``extend'' the fabric: on
$\Omega^4$ the corresponding system is \emph{infeasible} (960 constraints,
rank 214 --- corpus T-282), so a fourth mirror is not an engineering
challenge but a contradiction. Hamming checks the state; the rectangle layer
checks the \emph{law}.

There is a consequence that has no analogue in a layered machine, and it is
one of the genuinely new architectural facts \TALOS{} exposes.

\begin{principle}[Compute-diagnose identity]\label{prin:selfdiag}
Every tick both computes and emits its own syndrome. The cubic part of
$\LindOmega$ is generated by the line products (the closure law): a healthy
coherence triple \emph{closes} on its line, and a coherence that fails to
close is, by the same arithmetic, a non-zero syndrome. Error detection is
therefore not a separate pass over the state --- it is a byproduct of
advancing it. The machine cannot compute without simultaneously checking
itself.
\end{principle}

\noindent\emph{Why this is new.} In every prior architecture, checking is
extra work: a scrubber sweeps DRAM, a decoder runs after a read, a syndrome
is extracted by dedicated cycles. The layered design \emph{must} separate
compute from check because the code and the computation are different
structures. Under the collapse they are the same structure, so the check is
free in the strict sense that removing it would require removing the
computation. \TALOS{} is continuously self-diagnosing at no added cost ---
the DIAKRISIS self-diagnosis loop of the \SYNARC{} software layer is, at the
hardware level, simply reading the syndrome the tick already produced.

\subsection{Fault-tolerant composition}

Fault tolerance survives composition. Because any two lines of a projective
plane meet, the loss of any single sector leaves all $\binom{6}{2}=15$
surviving pairs still sharing a bus (verified for every removed sector), so a
holon degrades gracefully rather than partitioning. At the federation level,
the three-organism Golay grammar $[23,12,7]$ extends the same principle to
multi-fault tolerance across composed holons (the \SYNARC{} $\Omega$-organism
appendix); \TALOS{} inherits distance-3 protection at the single-holon level
and the federation inherits distance-7 protection above it, both from the
same projective family.

\begin{table}[H]
\centering\small
\caption{Fault tolerance: bolt-on codes versus code-as-topology.}
\label{tab:ecc}
\begin{tabular}{@{}p{3.6cm}p{3.4cm}p{4.7cm}@{}}
\toprule
\textbf{Scheme} & \textbf{Overhead} & \textbf{Coupling to computation} \\
\midrule
ECC-DRAM (SEC-DED) & $+12.5\%$ bits + decoder & separate; scrubbing is extra cycles \\
Surface code (quantum) & $\sim\!10^3$ phys./logical & separate; decodes against decoherence \\
\TALOS{} (Fano $=$ Hamming) & \textbf{zero} extra hardware & \textbf{identical}: the wiring is the check; every tick self-diagnoses (Principle~\ref{prin:selfdiag}) \\
\bottomrule
\end{tabular}
\end{table}
