\section{Energy: the machine pays for learning, not for computing}
\label{sec:energy}

\subsection{Where a von Neumann machine spends}

After Dennard scaling ended, the energy of a computation is dominated by
\emph{moving} its operands, not by the arithmetic: a floating-point operation
costs $\sim\!1$~pJ, an on-chip word $\sim\!1$~pJ, and an off-chip DRAM word
$\sim\!20$~pJ --- one to two orders more, and rising with distance. A
learning machine compounds this: training shuttles activations and gradients
between compute and memory on every step. The von~Neumann energy floor is a
\emph{data-movement} floor, and it sits some seven to eight orders of
magnitude above the thermodynamic (Landauer) limit~\cite{landauer1961,bennett1973} of the same computation.

\subsection{Where \TALOS{} spends: a new accounting}

The one instruction of \TALOS{} splits cleanly into a reversible part and two
entropy-exporting parts (Table~\ref{tab:three-terms}), and this split induces
an energy law with no counterpart in a layered machine.

\begin{claimbox}[Energy--learning law \status{T} for the reversible core, \status{C} for the coefficient]
The unitary term $-i[H_{\mathrm{eff}},\Gam]$ is an \emph{isometry} of every
monotone metric (T-262): it preserves the von~Neumann entropy of the state,
so by Landauer it costs \emph{zero} free energy. All free-energy expenditure
is charged to the two entropy-exporting terms:
\begin{equation}
\label{eq:energy-law}
E_{\mathrm{tick}} \;\geq\; kT\ln 2 \,\cdot\, \Delta H_{\mathrm{exported}},
\qquad
\Delta H_{\mathrm{exported}} \;=\; \underbrace{\Delta H_{\Regen}}_{\text{learning}}
\;+\; \underbrace{\Delta H_{\Diss}^{(+)}}_{\text{maintenance}},
\end{equation}
where $\Delta H_{\Regen}$ is the order the regeneration creates (the
free-energy-costly direction, $V_{\mathrm{hed}}=dP/d\tau>0$) and
$\Delta H_{\Diss}^{(+)}$ is only the entropy that must be re-exported to hold
the state against the dissipator's leak. \emph{Computing is free; only
forgetting-into-order (maintenance) and learning cost energy.}
\end{claimbox}

Two limiting cases make the law concrete.
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{A holon at its fixed point} ($\Gam=\rhostar$, converged, not
        learning) has $\Delta H_{\Regen}=0$; its only cost is the maintenance
        term, the free energy needed to hold $\rhostar$ against relaxation. If
        the task requires no forgetting, the dissipator can be quenched and
        the machine computes \emph{reversibly at the Landauer floor} --- net
        erasure zero.
  \item \textbf{A holon that is learning} spends in proportion to how fast it
        is moving toward its self-model, $\propto (dP/d\tau)_+$. Power tracks
        the learning rate directly: a machine that has learned draws less
        than a machine that is learning, with no change to what it computes.
\end{itemize}

\subsection{The numbers}

At $T=300$~K, $kT\ln 2 = 2.87\times10^{-21}$~J per irreversibly exported bit
($0.018$~eV). A learning-bounded tick that reshapes a few bits of the
$34$-parameter state costs on the order of $10^{-20}$~J; the worst case ---
re-exporting the entire state entropy every tick, which never happens away
from a cold start --- is $48\,kT\ln 2 = 1.4\times10^{-19}$~J. Against a single
off-chip memory access on a von~Neumann machine ($\sim\!2\times10^{-11}$~J),
the \TALOS{} learning-bounded tick sits roughly \textbf{nine orders of
magnitude} lower. The architecture does not \emph{reach} the Landauer floor
in any real substrate --- interface, control, and leakage overheads sit above
it --- but it \emph{removes the data-movement term that sets the
von~Neumann floor}, leaving a floor that is learning-limited rather than
transport-limited. That is the whole of the energy claim, and it is honest:
same computation, a floor nine orders lower, because there is nothing to move
and the reversible core is free.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=12cm, height=6.4cm,
  xlabel={computation performed (operations, arbitrary units)},
  ylabel={energy (log scale, J)},
  ymode=log, ymin=1e-21, ymax=1e-8,
  xmin=0, xmax=10, xtick=\empty,
  % legend sits in the empty mid-band between the (upper) von Neumann line and
  % the (lower) Talos/Landauer floors, so it covers no curve; short parallel
  % labels, the caption carries the full explanation.
  legend style={at={(0.5,0.52)}, anchor=center, font=\footnotesize,
                legend cell align=left, draw=gray!55, fill=white, rounded corners},
  grid=both, grid style={gray!15},
  every axis plot/.append style={very thick},
]
% von Neumann: energy rises ~linearly with computation (data movement)
\addplot[plosblue,domain=0.2:10,samples=40] {2e-11*x + 3e-11};
\addlegendentry{von Neumann floor ($\propto$ compute)}
% TALOS: flat, set by learning (here a converging learner: high early, decays)
\addplot[talosamber,domain=0.2:10,samples=60] {1.4e-19*exp(-0.45*x) + 6e-21};
\addlegendentry{\textsc{Talos} floor (learning-bounded)}
% Landauer floor
\addplot[plosgray,dashed,domain=0.2:10,samples=2] {2.87e-21};
\addlegendentry{Landauer floor $kT\ln2$}
\end{axis}
\end{tikzpicture}
\caption{Energy accounting (log scale, schematic with machine-checked
anchors). The von~Neumann floor rises with the amount of computation because
each operation moves data (blue). The \TALOS{} floor is set by the rate of
\emph{learning}: high during convergence, decaying --- as the holon converges
and stops learning (amber) --- toward the \emph{metabolic} floor of
\S\ref{subsec:metabolic}, which sits just above the Landauer line
(grey dashed) because a viable state must still be held against the
dissipator's leak.
Anchors: $kT\ln2=2.87\times10^{-21}$~J; DRAM word $\approx2\times10^{-11}$~J.}
\label{fig:energy}
\end{figure}

\begin{figure}[H]
\centering
\begin{tikzpicture}[font=\scriptsize, >=Stealth,
  box/.style={draw, rounded corners, thick, minimum height=8mm, align=center}]
  \node[box, fill=red!8, draw=red!60!black, minimum width=2.6cm] (port) at (0,0)
    {\textbf{power port}\\ negentropy in};
  \node[box, fill=plosblue!8, draw=plosblue!70!black, minimum width=3.0cm] (mach) at (4.1,0)
    {\textbf{the tick}\\ $\Gam$, Eq.~\ref{eq:lomega}};
  \node[box, fill=plosblue!14, minimum width=3.3cm] (work) at (9.4,1.35)
    {\textbf{work} (reversible)\\ $0$ J by Landauer};
  \node[box, fill=talosamber!14, minimum width=3.3cm] (maint) at (9.4,0)
    {\textbf{maintenance}\\ $P_{\mathrm{meta}} = kT\ln2\cdot\dot S_{\Diss}$};
  \node[box, fill=green!10, minimum width=3.3cm] (learn) at (9.4,-1.35)
    {\textbf{learning}\\ $kT\ln2\cdot\Delta H_{\Regen}$};
  \node[box, fill=gray!12, minimum width=1.9cm] (heat) at (13.3,-0.7) {heat\\ out};
  \draw[->, ultra thick, red!60!black] (port) -- node[above]{$r\,\rho_{\mathrm{env}}$} (mach);
  \draw[->, very thick, plosblue!70!black] (mach.east) to[out=25,in=180] (work.west);
  \draw[->, very thick, talosamber] (mach.east) -- (maint.west);
  \draw[->, very thick, green!45!black] (mach.east) to[out=-25,in=180] (learn.west);
  \draw[->, thick, plosgray] (maint.east) to[out=0,in=110] (heat.north);
  \draw[->, thick, plosgray] (learn.east) to[out=0,in=200] (heat.west);
  \node[plosgray, align=center] at (9.4,2.25) {\tiny the only branch a von~Neumann machine\\ \tiny cannot have: compute without paying};
  \draw[plosgray, ->] (9.4,2.0) -- (work.north);
\end{tikzpicture}
\caption{Where the joules go. All free energy enters through the power port
(\S\ref{subsec:powerport}). The reversible work branch is free (Landauer);
only holding order against the leak (maintenance, \S\ref{subsec:metabolic})
and creating order (learning) export entropy and cost. A layered machine has
no free branch: every operation moves data and pays.}
\label{fig:energyflow}
\end{figure}

\begin{mythbox}[The myth, continued --- ichor]
The giant's walking cost him nothing --- bronze does not tire. What the ichor
paid for was subtler: staying \emph{bronze} --- bright, whole, un-corroded
against the salt wind --- and getting \emph{better} at the patrol. That is
the energy law in one image: computation is the free walking; maintenance and
learning are the only two things the vein must pay for
(Eq.~\ref{eq:energy-law}).
\end{mythbox}

\begin{zerobox}[From zero --- why the toy needs feeding]
A wind-up toy runs down: not because moving is expensive, but because springs
leak. This machine's ``spring'' leaks too (the forgetting term), so it must
be fed --- that is the port. The surprise is what \emph{doesn't} cost: the
thinking itself. Feed it, and the thinking rides along free; starve it, and no
cleverness inside can keep it wound (\S\ref{subsec:powerport}). Its
food-bill tracks two things only: how much order it is \emph{holding}, and
how fast it is \emph{learning}.
\end{zerobox}

\subsection{The metabolic floor and the operating-point axis}
\label{subsec:metabolic}

The law above accounts for \emph{computation} (reversible, Landauer-free) and
\emph{learning} (the decaying $\Regen$ term). One term remains: the cost of
merely \emph{staying alive}. The dissipator $\Diss$ continually pushes $\Gam$
toward the maximally mixed heat-death state $I/7$; holding the state against
this leak is not free.

\begin{claimbox}[The metabolic floor \status{T}$+$\status{C} (T-273)]
At steady state a viable holon ($P>2/7$, $\Gam\neq I/7$) exports entropy at the
dissipator's strictly-positive rate $\dot S_{\Diss}>0$ (the entropy law
T-271). By Landauer, the minimum \emph{maintenance} power is
\[
  P_{\mathrm{meta}} \;\geq\; kT\ln 2\cdot \dot S_{\Diss} \;>\; 0,
\]
the \emph{active, irreversible} counterpart of the reversible energy floor. A
viable coherent machine's total power is therefore $0$ (reversible compute)
$+\ P_{\mathrm{meta}}$ (maintenance): it \textbf{cannot run for free}. Here
$\dot S_{\Diss}$ is the \emph{physical} entropy-production rate, so the floor is
\textbf{frequency-independent} (T-276): the tick-rate sets only an emulator's
real-time factor, not the fundamental cost. It is measurable, because
$\dot S_{\Diss}$ is a live telemetry observable (\S\ref{sec:platform}).
\end{claimbox}

The free energy that pays this floor has exactly one entry point: the
\textbf{power port} of \S\ref{subsec:powerport}. The drive
$\Lind_{\mathrm{feed}}$ imports structured negentropy from the environment;
regeneration \emph{directs} it into maintained coherence but cannot create it
(T-288) --- so an unplugged machine cannot pay $P_{\mathrm{meta}}$ and halts,
and a plugged one pays it out of the negentropy flux $r$ delivers. Maintenance
power, the ignition threshold $(r/\bar\gamma)_c$, and the halt latency after
severance are three faces of one budget line.

The floor scales with the \textbf{order maintained} (the distance $\ln 7 - S$
from $I/7$): a more coherent, more complex state is pushed harder by the
dissipator and costs more to hold --- \emph{the price of complexity}. This turns
the viability window into an explicit \textbf{design axis} rather than a single
set-point.

\begin{claimbox}[The operating-point axis \status{C} (T-274)]
The viable window $(2/7,3/7]$ is a capability--efficiency axis, not one point:
\begin{itemize}[leftmargin=1.4em,itemsep=1pt]
  \item \textbf{Lean edge $P\to 2/7$} --- efficiency-optimal: minimal maintenance
        power, minimal margin. The ``survival'' mode --- highest
        capability-per-watt (for capability $C=\Phi\times R$), least robust.
  \item \textbf{Rich edge $P\to 3/7$} (the T-124 attractor) ---
        capability-optimal: maximal $C=2/3$ and maximal dynamic range, at the
        highest maintenance power. The ``thriving'' mode.
\end{itemize}
The dynamics pull the machine toward $P=3/7$ by default; an engineer can run it
leaner toward $2/7$ to trade capability and margin for power. The optimal
set-point is \status{C} (it depends on the capability measure chosen); what is
\emph{robust} is the structure --- monotone cost across the window, two named
modes at its edges. Machine-verified (\texttt{t273\_metabolic.py},
\texttt{t274\_optimum.py}).
\end{claimbox}

\begin{claimbox}[The efficiency law \status{T}$+$\status{C} (T-276)]
Two refinements complete the thermodynamic budget. \textbf{(i)~Frequency-independence
\status{T}.} The metabolic floor is $P_{\mathrm{meta}} = kT\ln 2\cdot\dot S_{\Diss}$,
set by the \emph{physical} entropy-production rate: the per-tick entropy scales
with the step $\Delta\tau$, so the rate --- and the cost --- are independent of
the tick frequency. A faster clock buys real-time fidelity, not a lower floor.
\textbf{(ii)~Order-per-joule \status{T}.} Creating order (negentropy) costs
$\geq kT\ln 2$ per bit, so the regeneration's efficiency
$\eta = (\text{order created})/(\text{free energy spent}) \leq 1$ is
Landauer-bounded. \textbf{(iii)~Speed--efficiency tradeoff \status{C}.}
$\eta\to 1$ for \emph{quasi-static} regeneration and falls as the update
accelerates (irreversibility). The engineering rule follows: \textbf{run
regeneration as slowly as the dissipator leak permits} ($\kappa$ just above
$\gamma$) for maximal order-per-joule --- the coherent-machine analogue of a
slow, near-reversible heat engine. Machine-verified (\texttt{t276\_efficiency.py}).
\end{claimbox}
