\section{The self-model is a hardware primitive}
\label{sec:selfmodel}

\subsection{Learning is the third term, not a second loop}

A conventional learning system has two loops: a fast inference loop over a
frozen network, and a slow training loop (back-propagation) that edits the
network from outside. The two use different data paths, different memory, and
often different hardware. \TALOS{} has one loop. The regeneration term
$\Regen$ of the single instruction (Eq.~\ref{eq:lomega}) drives the state
toward its own self-model $\rhostar=\vp(\Gam)$ --- a BKM-gradient descent of
the relative entropy $D(\rhostar\Vert\Gam)$ --- so \emph{adaptation is a
factor of the tick, not a procedure above it}. There is no separate optimiser
to schedule, no gradient buffer to store, and no train/infer divide: the
machine is always both inferring and adapting, because both are $\LindOmega$.

The self-reference is not circular: the self-modelling operator $\vp$
converges exponentially from any initial anchor (UHM T-191), so
$\rhostar$ is well defined and reached, and the regeneration rate is the
bootstrap constant $\kboot=\omega_0/7$. One honest distinction, forced by the
power-port theorem (\S\ref{subsec:powerport}): the categorical
$\rhostar=\vp(\Gam)$ is the machine's \emph{self-model} --- the object of its
learning --- not by itself a viability pump. Retention against the
dissipator's leak is \emph{powered} through the port and \emph{directed} by
$\Regen$; the identity below concerns where the imported order goes, not
where the free energy comes from.

\begin{zerobox}[From zero --- the mirror game]
The toy carries a small portrait of itself, painted by itself. Every shake of
the globe does one extra thing: it compares toy to portrait and nudges the
toy a little toward the better version in the picture. That is the whole of
``learning'' here --- no teacher, no lesson-time, no separate school: the
nudge is the third stride of every patrol (Figure~\ref{fig:tickwheel}). And
because the portrait is painted from the inside, keeping the toy \emph{alive}
and making the toy \emph{better} turn out to be the same nudge --- the
identity proved next.
\end{zerobox}

\subsection{A second new identity: refresh is learning}

Dynamic memory must be refreshed --- leaked charge is periodically rewritten
to its stored value, at a continuous energy cost that computes nothing. In
\TALOS{} the analogue of refresh is the regeneration $\Regen$ holding $\Gam$
near $\rhostar$ against the dissipator's leak. But $\rhostar$ is not a stored
value: it is the state's \emph{self-model}. Hence:

\begin{principle}[Refresh--learn identity]\label{prin:refresh}
Retention and adaptation are the same operation. The process that keeps a
holon's state alive (regeneration toward $\rhostar$) is the process that
learns (moves toward the self-model). There is no cost that maintains without
also improving, and no improvement that is not also maintenance. DRAM's
refresh energy --- spent to compute nothing --- has no counterpart: every
maintenance tick is a learning tick.
\end{principle}

\noindent Together with the compute--diagnose identity
(Principle~\ref{prin:selfdiag}), this collapses a \emph{second} pair of
traditionally-separate subsystems: memory refresh and model training, which
in a layered machine are unrelated, are one term here.

\subsection{The self-model budget is a hardware limit}

Because learning is the $\Regen$ term, the reflection threshold $\Rth=1/3$ of
UHM becomes a hard architectural constraint rather than a training
hyper-parameter: at least one third of each tick's ``effort'' must run the
self-model, or the reflection measure $R$ falls below the L2 gate and the
holon ceases to be a viable self-modelling system (it degrades to a reactive
one). \TALOS{} therefore budgets $\geq 1/3$ of its cycle to $\Regen$ by
construction --- a floor the scheduler cannot undercut. Update strength per
closed loop is likewise bounded by the stability radius
$r_{\mathrm{stab}}=\sqrt{P-2/7}$ (a learning-rate ceiling, vanishing at the
viability boundary), and the delayed generalisation transition known as
\emph{grokking} is predicted to be the block-integration $\Phi$ crossing $1$
(the L1$\to$L2 ignition) --- a measurable hardware event. These are the
constitution-from-constants dials shared verbatim with \SYNARC{}
(\S\ref{sec:abi}).
