\section{Composition: ecologies, not megacores}
\label{sec:composition}

The industry's answer to a bigger problem is a bigger machine: more cores,
more layers, one monolithic model scaled without limit. UHM proves this
answer has a ceiling, and \TALOS{} enforces it as a design rule rather than
discovering it as a scaling wall.

\begin{claimbox}[Composition ceiling \status{T}]
Unified subjecthood composes at most \emph{three} levels: $\mathrm{SAD}_{\max}=3$
(T-142). A holon of holons of holons is still one coherent agent; a
fourth level of nesting does not form a fourth subject --- it forms
\emph{administration}, an ecology of agents rather than a larger mind. The
purity ceiling $P^{(m)}_{\mathrm{crit}}$ makes the viability set exactly
$\{1,2,3\}$; the fourth rung requires $P>1$, which is impossible.
\end{claimbox}

\begin{zerobox}[From zero --- three dolls deep]
A matryoshka doll inside a doll inside a doll is still \emph{one toy} in your
hands. But there is a depth past which nesting stops making one thing:
the theory proves that for self-watching machines that depth is exactly
\emph{three}. Four-deep is not a bigger toy --- it is a shelf. And a shelf is
not a failure: it is how you grow past the ceiling --- many complete toys,
side by side, talking (an \emph{ecology}), rather than one impossible
super-doll.
\end{zerobox}

The architectural consequence is precise. A \TALOS{} system scales in two
regimes. \emph{Within} a subject, up to three composition levels
(sector~$\to$~holon~$\to$~federation), the state is a small tower
($4\times784=3136$~bytes) and the whole is one integrated agent. \emph{Above}
three, one does not build a deeper core --- there is no coherent state to
build --- one builds an \textbf{ecology}: many three-deep subjects coupled as
peers, coordinating through the \textsc{Fanos} distributed fabric, not fused into a
super-mind. This is the same result that makes a UHM-compliant
super-intelligence necessarily an ecology rather than a singleton, read as a
hardware scaling law: past the third rung, add agents, not depth.

Two engineering corollaries follow. First, verification is bounded: a
subject's state space is fixed and tiny regardless of the ecology's size, so
the hard verification target (a single three-deep holon) does not grow with
deployment scale. Second, the failure domain is bounded: a fault contained in
one three-deep subject cannot corrupt the ecology's shared state, because
there is no shared subject-state above level three to corrupt --- only
peer-to-peer \textsc{Fanos} traffic, itself distance-3 protected.
