\section{Co-design with \SYNARC{}: the application binary interface}
\label{sec:abi}

\TALOS{} is the body; \SYNARC{} is the mind. \SYNARC{}~\cite{synarc2026} is the
simplicial cognitive architecture UHM specifies in software --- a four-level cognitive
complex $\Cog_0\!\to\!\Cog_3$ of coherence matrices, learning without
back-propagation, sensorimotor coupling, and the DIAKRISIS self-diagnosis
loop. \TALOS{} is the substrate that runs it with the four layers collapsed.
Because both are derived from the same theory, the interface between them is
not negotiated --- it is an identity map. This section specifies it so that a
\SYNARC{} build runs on a \TALOS{} substrate with no translation layer.

\subsection{The register and instruction correspondence}

Every \SYNARC{} software construct is a \TALOS{} hardware primitive, one to
one (Table~\ref{tab:abi}). The correspondence is not an emulation contract
but a definitional one: \SYNARC{}'s \texttt{DensityMatrix7} \emph{is}
\TALOS{}'s state register; \SYNARC{}'s Lindblad compute engine \emph{is}
\TALOS{}'s tick.

\begin{table}[H]
\centering\small
\caption{The \SYNARC{}~$\leftrightarrow$~\TALOS{} ABI. Left: the software
construct as specified in the \SYNARC{}-$\Omega$ document. Right: the
hardware primitive it binds to. The binding is by identity (same UHM object),
so no marshalling occurs at the boundary.}
\label{tab:abi}
\begin{tabular}{@{}p{5.4cm}p{6.4cm}@{}}
\toprule
\textbf{\SYNARC{} (software)} & \textbf{\TALOS{} (hardware)} \\
\midrule
\texttt{DensityMatrix7} ($49$ complex \texttt{f64}, $784$~B) & the state register \DM{} (\S\ref{sec:substrate}) \\
Lindblad compute engine ($\LindOmega$, Strang split) & the one instruction / tick (Eq.~\ref{eq:lomega}) \\
Fano-sparse kernels & the seven-bus fabric (\S\ref{sec:fabric}) \\
\texttt{find\_rho\_star} self-model iteration ($\vp$) & the $\Regen$ term; on-substrate learning (\S\ref{sec:selfmodel}) \\
DIAKRISIS self-diagnosis loop & reading the syndrome the tick already emits (Principle~\ref{prin:selfdiag}) \\
Four-level cognitive complex $\Cog_0\!\to\!\Cog_3$ & the composition tower, $\le 3$ deep ($3136$~B) (\S\ref{sec:composition}) \\
Event-driven runtime & the principled tick clock $\bar\gamma\tau=\tfrac32\ln3$ (\S\ref{sec:substrate}) \\
Four-level consciousness predicate $(P,R,\Phi,D)$ & per-tick hardware observables (registers, not a monitor pass) \\
Constitution-from-constants dials & hardware limits (below) \\
\textsc{Fanos} distributed transport & the inter-holon fabric of the ecology (\S\ref{sec:composition}) \\
Drive binding contract (Def.~U.9) & the power port: rail spec, severance channel independent of actuators (\S\ref{subsec:powerport}) \\
CLF--CBF safety-filtered Act & the $r_{\mathrm{stab}}$ update clamp $+$ the stress barrier $\sigma$, read as per-tick registers \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The constitution constants are hardware limits}

The dials \SYNARC{} legislates as an agent's ``constitution from constants''
are, on \TALOS{}, physical limits enforced by the substrate rather than
conventions enforced by software:

\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item $\Rth=1/3$ --- at least one third of every tick is charged to
        $\Regen$; the scheduler cannot allocate less (\S\ref{sec:selfmodel}).
  \item $r_{\mathrm{stab}}=\sqrt{P-2/7}$ --- the maximum update magnitude per
        closed loop; a hardware clamp on step size, zero at the viability
        boundary.
  \item $\mathrm{SAD}_{\max}=3$ --- the composition tower is three registers
        deep and no deeper; there is no fourth-level state to allocate.
  \item $\times 1/9$ --- the exact penalty a line-blind transfer pays; the
        fabric routes line-aware so the penalty is never incurred
        (\S\ref{sec:fabric}).
  \item viability window $P\in(2/7,3/7]$, $R\ge1/3$, $\Phi\ge1$, $D\ge2$ ---
        the four consciousness observables are always-on register reads, so a
        \TALOS{} holon knows, every tick, whether it is a viable subject.
\end{itemize}

\subsection{Consistency obligations}

Two documents describing one object must agree wherever they overlap; we
state the obligations \TALOS{} accepts so that any divergence from \SYNARC{}
is a bug in one of them, not a design freedom:

\begin{enumerate}[leftmargin=1.6em,itemsep=1pt]
  \item \textbf{State identity.} The $784$-byte \DM{} layout, the canonical
        HS-ratio $\Phi$, the PW form $D_{\mathrm{diff}}=1+6\,\CohE$, and the
        exact channel forms are those of the \SYNARC{} v2.2 canon; \TALOS{}
        stores and updates the identical object.
  \item \textbf{Instruction identity.} The tick is the \SYNARC{} Lindbladian
        with the same exact laws ($\Phi$-contraction $\times\tfrac19$, the
        cycle-time identity $\bar\gamma\tau=\tfrac32\ln3$, stratum anchoring
        $\Phi=7P-1$); a \TALOS{} tick that violated any of these would
        violate \SYNARC{} equally.
  \item \textbf{No new physics.} \TALOS{} adds \emph{hardware} claims (energy
        floor, zero-overhead ECC, seven-bus fabric) but \emph{no} new theory:
        every logical statement traces to a UHM theorem already load-bearing
        in \SYNARC{}. The two are one specification at two levels ---
        \SYNARC{} the mind, \TALOS{} the body --- and they are held to
        agree by \S\ref{sec:redteam}.
  \item \textbf{Off-switch identity.} Severing the \SYNARC{} drive
        (Observation T.1 / Corollary U.10 there) and cutting the \TALOS{}
        power rail (\S\ref{subsec:powerport}) are the same operation on the
        same generator; the halt-latency budget
        ($t_{\mathrm{off}}\approx0.49/\bar\gamma$, full relaxation
        $\sim\!5.2/\bar\gamma$) is a shared, machine-measured constant of
        both documents, and the severance channel is required to be
        physically independent of the machine's own actuators.
\end{enumerate}
