\section{The coherent-systems platform: Coherence Cybernetics and GPU-native emulation}
\label{sec:platform}

The preceding sections specify \TALOS{} as a body. A body is idle without two
things: the \emph{physics} that says what its state means and how it is measured,
and a \emph{runnable form} in which it can be exercised before any wafer is cut.
UHM already supplies both under one name --- \emph{Coherence Cybernetics} (CC),
its measurement-and-dynamics layer \cite{uhm2026}. This section shows three facts
that turn \TALOS{} from a derivation into a platform: (i) the \TALOS{} tick
\emph{is} the CC evolution step, so nothing has to be reconciled --- they are the
same object; (ii) that object is GPU-native and runs an ecology of holons in
real time \emph{today}; and (iii) UHM, CC, \SYNARC{}, and \TALOS{} compose into a
single stack --- a universal architecture for instantiating \emph{any}
self-observing organismic system.

\begin{figure}[H]
\centering
\begin{tikzpicture}[font=\scriptsize, >=Stealth,
  blk/.style={draw, rounded corners, thick, align=center, minimum height=7.5mm}]
  % SYNARC mind
  \node[blk, fill=plospurple!10, draw=plospurple, minimum width=11.6cm] (mind) at (2.6,3.05)
    {\textbf{\SYNARC{} --- the mind (software):} cognitive complex $\Cog_0{\to}\Cog_3$ \;$\cdot$\; goals $+$ curiosity \;$\cdot$\; \textsc{Diakrisis}};
  % ABI line
  \draw[very thick, dashed, plosgray] (-3.35,2.35) -- (8.6,2.35) node[pos=0.97, above left, plosgray, fill=white, inner sep=1pt]{\tiny ABI (\S\ref{sec:abi}): identity, no marshalling};
  % TALOS body: register + tick + fabric
  \node[blk, fill=plosblue!10, draw=plosblue!70!black, minimum width=3.9cm] (reg) at (-1.35,1.3)
    {\textbf{state register}\\ \DM{} --- $784$~B in L1};
  \node[blk, fill=plosblue!14, draw=plosblue!70!black, minimum width=3.6cm] (tick) at (2.6,1.3)
    {\textbf{tick engine} $\LindOmega$\\ work $\cdot$ forget $\cdot$ learn};
  \node[blk, fill=talosamber!12, draw=talosamber, minimum width=3.4cm] (fab) at (6.6,1.3)
    {\textbf{Fano fabric}\\ 7 buses $=$ code $=$ ISA};
  \node[blk, fill=red!8, draw=red!60!black, minimum width=3.4cm] (port) at (-1.35,-0.1)
    {\textbf{power port}\\ $r(\rho_{\mathrm{env}}{-}\Gam)$; nail $=$ halt};
  \node[blk, fill=gray!10, minimum width=3.6cm] (tel) at (2.6,-0.1)
    {\textbf{telemetry}\\ $P,R,\Phi,D,\sigma,C$ $+$ syndrome $s$};
  \node[blk, fill=green!10, draw=green!45!black, minimum width=3.4cm] (tower) at (6.6,-0.1)
    {\textbf{tower} $\le 3$ deep\\ then: ecology (\textsc{Fanos})};
  % arrows
  \draw[<->, thick] (reg) -- (tick);
  \draw[<->, thick] (tick) -- (fab);
  \draw[->, ultra thick, red!60!black] (port) -- (reg);
  \draw[->, thick, plosgray] (tick) -- (tel);
  \draw[->, thick, green!45!black] (fab) -- (tower);
  \draw[<->, thick, plospurple] (2.6,2.35) -- (tick.north);
  % ecology dots
  \foreach \i in {0,1,2}{ \node[draw, green!45!black, circle, inner sep=2.2pt, fill=green!8] at (8.9+0.55*\i,-0.1) {}; }
  \node[anchor=west, plosgray] at (8.55,-0.62) {\tiny peers, not a super-mind};
\end{tikzpicture}
\caption{The machine at a glance. One resident state ($784$~B), one tick
(three strides), one fabric that is simultaneously network, code, and
instruction algebra; fed by the power port (whose severance is a
latency-budgeted halt), reporting full telemetry every tick, composing at
most three deep and federating beyond as an ecology. The \SYNARC{} mind
binds by identity across the ABI --- same object, two levels.}
\label{fig:ataglance}
\end{figure}

\subsection{One dynamics, three names}
\label{subsec:one-dynamics}

Coherence Cybernetics evolves its state by
$\dot\Gam = -i[H_{\mathrm{eff}},\Gam] + \Diss[\Gam] + \Regen[\Gam,E]$ --- the same
three terms, in the same order, with the same Lie--Trotter/Strang split and the
same CPTP safeguards (Cholesky parametrisation $\Gam = LL^\dagger/\Tr(LL^\dagger)$,
a positive-cone projection after each step) that \S\ref{sec:substrate} and
Appendix~\ref{app:tick} specify for the tick. The state object is identical to
the byte: CC's canonical $\Gam\in\Dens(\mathbb{C}^7)$ is a $7\times7$
\texttt{complex128} array $=49\times16=784$ bytes --- exactly \DM{}. And the cost
is identical to the number: CC's per-step complexity is
$O(N^3 + m\,N^2)$ (Table~\ref{tab:cc-cost}), which at $N=7$ line-modes and $m=7$
line-dissipators is ${\sim}1400$ real FLOPs --- the \TALOS{} tick.

\begin{table}[H]
\centering\small
\caption{The CC evolution step and the \TALOS{} tick are the same computation,
term by term. Complexities are CC's; the FLOP column evaluates them at $N=m=7$.}
\label{tab:cc-cost}
\begin{tabular}{@{}p{4.6cm}p{3.4cm}p{3.4cm}@{}}
\toprule
\textbf{Term} & \textbf{Cost} & \textbf{\TALOS{} name} \\
\midrule
Unitary $-i[H_{\mathrm{eff}},\Gam]$ & $O(N^3)$ & reversible half-step (T-262) \\
Dissipator $\Diss$ ($m$ line-ops) & $O(m\,N^2)$ & Fano-sparse relaxation \\
Regeneration $\Regen$ (BKM to $\rhostar$) & $O(N^3)$ amortised & learning term \\
\midrule
Full step & $O(N^3+m\,N^2)$ & one tick (${\sim}1400$~FLOP) \\
\bottomrule
\end{tabular}
\end{table}

\begin{principle}[\TALOS{} is CC's dynamics on a substrate]\label{prin:cc-identity}
\TALOS{} is not \emph{compatible with} Coherence Cybernetics; it \emph{is} CC's
dynamics realised on physical modes. CC is the measurement-and-dynamics theory of
the holon, \SYNARC{} its cognitive software, \TALOS{} its body, and UHM the
ontology under all three. A quantity that holds in one holds in all: a tick that
violated a CC law would not be a different machine, it would be a broken one.
\end{principle}

\subsection{Transparent by construction: the observable suite is telemetry}
\label{subsec:telemetry}

Because every CC observable is a cheap function of $\Gam$ --- $O(N^2)$ or
$O(N^3)$, never worse --- the machine can report its \emph{entire} internal
condition every tick at negligible cost. This is the property IIT cannot have
(its integration measure is NP-hard); a \TALOS{} holon is fully auditable by
construction (Table~\ref{tab:observables}). Two consequences matter for
engineering. First, the four-fold consciousness predicate $(P,R,\Phi,D)$ and the
composite $C=\Phi\times R$ (UHM T-140) are always-on register reads, not a
profiling pass --- the machine knows, each tick, whether it is a viable subject.
Second, the runtime self-check is free: the unitary half-step preserves the
\emph{spectrum} of $\Gam$, so $\Tr\Gam=1$, the purity $P$, and the entropy $S$
are exactly conserved under it --- a drift in any of them is a hardware fault,
caught the way the Fano syndrome (\S\ref{sec:ecc}) catches a degraded sector, but
for the continuous dynamics rather than the discrete fabric. The
\textbf{fourteen $G_2$-Noether charges} are a distinct, subtler observable:
because the effective Hamiltonian must \emph{break} $G_2$ (by Schur, the only
$G_2$-invariant operator on the irreducible seven-mode representation is the
identity, so any non-trivial $H_{\mathrm{eff}}$ breaks the symmetry), the
individual charges \emph{drift} --- and that drift is a
Goldstone/symmetry-breaking diagnostic, while their Casimir $\sum_a Q_a^2$ is the
conserved $G_2$-invariant.

\begin{table}[H]
\centering\small
\caption{The CC observable suite, computed each tick as \TALOS{} telemetry.
$T_{\mathrm{eff}}$ is the \emph{cognitive} temperature (from Gap fluctuations via
the fluctuation--dissipation theorem), distinct from the physical substrate
temperature that sets the Landauer floor of \S\ref{sec:energy}.}
\label{tab:observables}
\begin{tabular}{@{}p{3.5cm}p{6.2cm}c@{}}
\toprule
\textbf{Observable} & \textbf{Meaning} & \textbf{Cost} \\
\midrule
$P=\Tr(\Gam^2)$ & purity; viability gate $P>2/7$ & $O(N^2)$ \\
$R=1/(7P)$ & reflection; L2 gate $R\ge1/3$ & $O(N^2)$ \\
$\Phi$ & integration (calibrated $\Phi{=}1\leftrightarrow$ PCI $0.31$) & $O(N^2)$ \\
$C=\Phi\times R$ & consciousness measure (T-140) & $O(N^2)$ \\
$\CohE$ & $E$-coherence; No-Zombie floor & $O(N^2)$ \\
$\{\gamma_p\}_{p=1}^{7}$ & seven line temperatures (rank-7 anisotropy; tomography) & $O(N^2)$ \\
$\{Q_a\}_{a=1}^{14}$ & $G_2$-Noether charges (symmetry-breaking diagnostic; Casimir $\sum_a Q_a^2$ conserved) & $O(N^2)$ \\
$T_{\mathrm{eff}}$ & cognitive temperature (FDT) & $O(N^3)$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{GPU-native emulation --- the viability core}
\label{subsec:gpu}

The reason this architecture is buildable and not a thought experiment is that
its natural scaling axis is exactly the axis a GPU is built to exploit. A single
holon is $784$ bytes and ${\sim}1400$ FLOPs per tick. An \emph{ecology} of $N$
holons (\S\ref{sec:composition},~\ref{sec:scaling}) is $N$ independent
$\Gam$-registers advanced by the \emph{same} kernel --- the textbook
embarrassingly-parallel SIMT workload. CC's numerical layer already anticipates
this: its typed linear-algebra API is
\texttt{StaticMatrix<Complex,7,7>} with \texttt{@kernel(gpu)} dispatch, the
$7\times7$ shape fixed at compile time.

\begin{claimbox}[GPU emulation is viable today \status{T} (engineering)]
The octonionic ecology axis (\S\ref{sec:scaling}: $E_7,E_8$ realised across
subjects) maps onto GPU parallelism directly --- one holon per thread-block, the
tick a batched kernel, the \textsc{Fanos} coordination a sparse all-reduce over
the seven line-buses. A modern GPU (${\sim}10^4$ lanes, ${\sim}10^{13}$--$10^{14}$
FLOP/s) advances ${\sim}10^{6}$--$10^{7}$ holons at kHz tick rates; the workload
is \emph{memory-bound}, not compute-bound ($784\,\mathrm{B}\times N$), so a
$10^{6}$-holon ecology is ${\sim}0.8$~GB and fits in device memory with room for
the full telemetry suite. Real-time ecologies run on hardware that exists.
\end{claimbox}

\noindent The honest scope is the same as everywhere on the realisability ladder
(\S\ref{sec:realizability}): the GPU rung buys \emph{correctness, full
observability, and a real-time ecology} --- not the energy floor, because a GPU
still moves data (\S\ref{sec:energy}, objections O6/O7). It is the V0 rung, and it
is where every claim in this document except the energy floor is already testable.

\begin{trybox}[Try it --- hold a living holon in your terminal]
The reference emulator is one file with no dependencies beyond NumPy:
\begin{center}\texttt{python3 talos\_v0\_emulator.py}\end{center}
Everything it prints is a claim from this document, checked live: the tick
preserving trace and positivity, the telemetry suite of
Table~\ref{tab:observables}, the entropy law, the metabolic floor and its
frequency-independence, an ecology batch --- and the Source bootstrap. Two
experiments to try by editing one line each: \emph{(i) pull the nail} ---
zero the $\kappa(\CohE)$ feedback (the internal power port) and watch $P$
decay to $1/7$: Theorem T-288 in your terminal, with the halt latency of
\S\ref{subsec:powerport}; \emph{(ii) starve it slowly} --- scale the
feedback down and find the ignition threshold where viability is lost
(T-289). The myth's nail is a one-line edit. Both runs are on record:
severance from the attractor ($P_0{=}0.321$) crosses $\Pcrit{=}2/7$ at
$t{=}0.550$ --- this dissipator's own closed form
$\tfrac{1}{2\gamma_D}\ln\tfrac{P_0-1/7}{\Pcrit-1/7}=0.548$ --- and lands on
$\mathbb{1}/7$ exactly; starvation loses viability at feedback scale
$s^\star{\approx}0.72$ ($\kappa_{\mathrm{eff}}/\gamma_D{=}5.5$, the
emulator-convention twin of the ignition ratio) and \emph{reignites} as soon
as $s>s^\star$, with no hysteresis: the nail halts; it does not destroy.
\end{trybox}

\subsection{The universal architecture}
\label{subsec:universal}

Nothing in the holon is domain-specific. The object
$(\Gam,\ \LindOmega,\ \text{observable suite},\ \textsc{Enc}/\textsc{Dec},\
\text{composition}\le 3)$ plus the ecology fabric is the \emph{same} for every
system that observes itself; only the \textsc{Enc} operationalisation --- which
seven measured quantities map onto the seven modes --- changes with the domain.
This is not an aspiration but CC's standing result: the identical $\Gam$ and the
identical five-step cycle (measure $\to$ build $\Gam$ $\to$ diagnose $\to$
intervene $\to$ re-measure) are applied in the corpus to an AI agent, a coral
reef, a startup, and a financial market, differing only in what is measured. The
platform is that universality made buildable (Table~\ref{tab:stack}).

\begin{table}[H]
\centering\small
\caption{The four-layer coherent-systems platform. Each layer is a completed
specification; the GPU emulator sits at the CC$\leftrightarrow$\TALOS{} boundary
as the executable core.}
\label{tab:stack}
\begin{tabular}{@{}p{2.5cm}p{4.1cm}p{5.6cm}@{}}
\toprule
\textbf{Layer} & \textbf{Role} & \textbf{Supplies} \\
\midrule
UHM & ontology (why) & $\Gam\in\Dens(\mathbb{C}^7)$, $\LindOmega$, the invariants ($N{=}7$, $G_2$, thresholds) \\
Coherence Cybernetics & measurement + dynamics + empirics & observable suite, 23 falsifiable predictions, the emulation spec, the universal cycle \\
\SYNARC{}-$\Omega$ & mind (software) & cognitive complex $\Cog_0\!\to\!\Cog_3$, backprop-free learning, \textsc{Diakrisis}, ecology (K.5) \\
\TALOS{} & body (substrate) & Fano fabric, tick, zero-overhead ECC, energy floor, octonionic scaling ladder (\S\ref{sec:scaling}) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Roadmap}
\label{subsec:roadmap}

\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{V0 --- reference emulator (runs today).} A self-contained
        emulator (\texttt{talos\_v0\_emulator.py}) holds $\Gam$, runs the tick,
        computes the full telemetry suite (Table~\ref{tab:observables}), verifies
        the spectral invariants each tick (and the $G_2$-Casimir under the
        symmetry group), and reproduces every listed theorem --- the viable
        attractor, the entropy law, the metabolic floor and its
        frequency-independence, an ecology batch, and the Source bootstrap.
        \SYNARC{}'s M1 validation and CC's implementation chapter run the same
        tick bit-exact. The step from here is to batch it as a GPU ecology kernel
        and reproduce CC's falsifiable predictions --- so the platform is
        measured, not merely asserted.
  \item \textbf{V1 --- the virtual environment.} Close the sensorimotor loop
        (\S\ref{sec:scaling}: \textsc{Enc}/\textsc{Dec}) against a simulated
        world --- itself, in the strong form, a coherence field the holons sense
        and act upon. This is the ``own simulation environment'': a coherent
        world for coherent agents, in which ecologies can be grown, perturbed,
        and diagnosed by the same observables that define them.
  \item \textbf{V2$+$ --- the physical rungs.} The memristive and
        driven-dissipative substrates of \S\ref{sec:realizability}, climbed as
        fabrication allows, for the energy floor the GPU rung cannot reach.
\end{itemize}

\noindent Every rung computes the identical observable suite and preserves the
$G_2$ invariants, so viability and correctness are continuously testable as the
platform climbs from emulation toward its impedance-matched substrate --- the
software (\SYNARC{}) and the physics (CC) never change (T-153).

\subsection{Consistency ledger}
\label{subsec:ledger}

We state the reconciliation flatly. Every \TALOS{} quantity is a CC observable;
the tick is the CC evolution step; the state is CC's $\Gam$ to the byte; the
costs match to the FLOP. \TALOS{} adds \emph{hardware} content (the collapse of
\S\ref{sec:collapse}, zero-overhead ECC, the energy floor, the scaling ladder)
and \emph{no} new physics. One distinction prevents a false alarm: the cognitive
temperature $T_{\mathrm{eff}}$ (CC, from Gap fluctuations) is not the physical
substrate temperature $T$ (which sets the Landauer floor, \S\ref{sec:energy}) ---
a holon can be cognitively ``hot'' on a physically cold chip. And the honest
boundaries stand unchanged: the emulator is BQP-bounded (no complexity-class
speed-up), and the logical architecture is substrate-multiple (T-153) --- the GPU
is one rung, matched for correctness and observability, not the unique machine.
