\section{The energy floor, derived}
\label{app:energy}

We derive Eq.~\eqref{eq:energy-law}. Landauer's principle states that erasing
one bit of information --- reducing the state's entropy by $\ln 2$ nats while
exporting it to a bath at temperature $T$ --- costs at least $kT\ln 2$ of free
energy. The converse (creating order) costs the same. The free-energy budget
of a tick is therefore $kT\ln 2$ times the number of bits of entropy the tick
\emph{exports}.

\paragraph{The three terms.} Decompose the tick (Eq.~\ref{eq:lomega}).
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{Unitary} $-i[H_{\mathrm{eff}},\Gam]$: by T-262 this is an
        isometry of every monotone metric and preserves the von~Neumann
        entropy $H(\Gam)=-\Tr(\Gam\ln\Gam)$ exactly, $\dot H_{\text{unitary}}=0$.
        It exports no entropy, hence costs no free energy. This is the
        reversible core.
  \item \textbf{Dissipator} $\Diss$: relaxation toward $\mathbb{1}/7$
        \emph{raises} $H$ (loses information, thermodynamically downhill,
        heat out). Holding the state against this leak requires re-exporting
        $\Delta H_{\Diss}^{(+)}$ per tick --- the maintenance term.
  \item \textbf{Regeneration} $\Regen$: descent toward $\rhostar$
        \emph{lowers} $H$ (creates order, the self-model), which by Landauer
        costs $kT\ln2\cdot\Delta H_{\Regen}$ of free energy per tick. This is
        the learning term, and it is exactly the ``regeneration paid in free
        energy'' bookkeeping of UHM's cosmological accounting
        ($V_{\mathrm{hed}}=dP/d\tau$).
\end{itemize}
Summing, $E_{\mathrm{tick}}\ge kT\ln2\,(\Delta H_{\Regen}+\Delta H_{\Diss}^{(+)})$,
Eq.~\eqref{eq:energy-law}. At the fixed point $\Delta H_{\Regen}=0$ (nothing
to learn) and, if the task tolerates no forgetting, $\Diss$ can be quenched,
leaving a reversible tick at zero net erasure --- the Landauer floor.

\paragraph{Numbers ($T=300$~K).} $kT\ln2=1.380649\!\times\!10^{-23}\cdot300\cdot\ln2
=2.87\times10^{-21}$~J $=0.018$~eV per bit. Worst-case tick (re-export the
entire $48$-DOF state; $34$ physical after the $G_2$ gauge): $48\,kT\ln2=1.38\times10^{-19}$~J. A learning-bounded
tick reshaping $\sim\!4$ bits: $1.1\times10^{-20}$~J. Von~Neumann off-chip
word: $\sim\!2\times10^{-11}$~J. Ratio (DRAM word $/$ learning-bounded tick):
$\sim\!1.7\times10^{9}$. Hence the $\sim$nine-orders headline
(\S\ref{sec:energy}). All values reproduced by the verifier
\texttt{talos\_energy.py}. The floor is a ceiling on efficiency, not a
promise a real device meets it; the architectural content is that the
data-movement term --- which sets the von~Neumann floor --- is absent.
